WFH and broadband speed (title needs rework)

### 2020-12-11 paper/v1/broadband.speed.covid.tex
updates in text, tables, figures

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

During the pandemic, working from home using digital technologies,
whether partially or exclusively, was transformed from a niche means of
accessing work, albeit one that had been on a slow, upward trend, to a
widespread way of life in many countries. The ability to work from home
or telecommute meant millions retained their jobs and, to a varying
extent, maintained productivity during periods of strict lockdown around
the world. However, this ability has not been evenly distributed
socially or spatially, creating a new type of digital divide. On one
side are those who can work from home, supported by digital
technologies, and have thus been able to enjoy both economic resilience
and greater personal safety. On the other side, previously employed
individuals have been forced to accept furlough or redundancy packages
unless they are part of the cadre of essential workers, who are
potentially at high risk of infection. Whilst the basis for this new
digital divide has been viewed as mainly occupational, here we consider
whether the divide is also technological.

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds, may reinforce or redress the spatial
and social dimensions of the digital division exposed by the pandemic.
To do so, we employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. \textbf{need to add
urban-rural to analysis} Our analysis enables us to better understand
how the spatial and social distribution of both occupations and online
accessibility intersect to enable or hinder the practice of
telecommuting at a time of extreme demand. We will also consider what
lessons can be learned from this time for a future where telecommuting
is likely to remain a common means of accessing work, and broadband
services and infrastructure must be fit for purpose. \textbf{LET'S LEAVE
IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ}

@@

The structure of this paper is as follows. First we review the
literature on telecommuting and digital divides to better understand the
origins of the new digital divide revealed by the pandemic and its
impact on the economic resilience of different places. We then describe
our data and methodology. Our results section first offers
classification of how internet services vary across the UK local
authorities and then assesses whether these clusters replicate or
repudiate other socio-economic and geographic patterns of economic
resilience.

\hypertarget{literature-review}{%
\section{Literature review}\label{literature-review}}

\hypertarget{from-telecommuting-to-wfh}{%
\subsection{From telecommuting to
\#WFH}\label{from-telecommuting-to-wfh}}

In this analysis, the terms `telecommuting' and `working from home' are
used interchangeably, as most remote labour during the Covid-19 crisis
was carried out in the homes of individual employees rather than any
other location \citep{eurofound2020}. However, it should be noted that
previous research has explored how telecommuting can occur in other
places, including satellite offices or on public transport
\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
also used a variety of definitions to measure the level of telecommuting
within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or
part-time, and those who use digital technologies to work remotely
full-days or part-days
\citep{allen2015effective, bailey2002review, haddad2009examination}. No
matter the definition, the option and capability to telecommute or work
from home has never been equally distributed spatially or
socio-economically any more than different industries and employment
opportunities have. For example, studies from the United States, the
Netherlands, and the UK indicate that telecommuters are most likely to
hold professional, managerial, and technical occupations, where the
workforce is better educated and wealthier, and that there is suppressed
demand among women and part-time workers
\citep{headicar2016move, peters2004employees, singh2013modeling}.
Opportunities for working from home during the current pandemic have
likewise not been equally spread across the workforce.
\citet{NBERw26948} indicated that in the US, managers, educators, as
well as those working in computer-related occupations, finance, and law
can easily work from home, and that occupations with opportunities to
telecommute are associated with higher earnings. This is not the case
for the workforce occupied in more spatially fixed occupations, from
farming, construction and manufacturing to hospitality and care
services. In the US, these occupations tend to be lower-income,
non-white, without a university degree, live in rental accommodation and
lack health insurance \citep{NBERw27085}. Similar trends can be observed
for other countries. For example, \(75\)\% of workers with tertiary
education worked from home in Europe during spring \(2020\), whilst only
\(34\)\% of workers with secondary education and \(14\)\% of those
primary education did so \citep{eurofound2020}.

\hypertarget{digital-divides-and-economic-resilience}{%
\subsection{Digital divides and economic
resilience}\label{digital-divides-and-economic-resilience}}

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
These returns have been even greater during the Covid-19 crisis, when
those with the ability to telecommute also have the ability to maintain
their employment whilst protecting their health. However, the success of
these arrangements has been dependent upon the first level digital
divide, which is associated with access and quality of internet
connectivity. For example, the systematic review from
\citet{SALEMINK2017360} highlights the infrastructure quality
differences between urban and rural areas in various advanced economies.
Whether this variation in infrastructure quality affects the spatial
footprint of telecommuting has not previously been measured, although
there are indications that those who purchase high speed connections
consume more data of all sorts and use their connections for a variety
of purposes \citep{hauge2011consumer}. Another study identifies a
correlation between access to internet services and a reduction in
household transport spend across 33 countries \citep{bris2017ict}.
Whether the implication is less travel because of increased
telecommuting, or whether internet access enables more efficient travel,
this finding is an indication of the potential household savings better
internet services offer. Thus, the extreme demand during the pandemic
provides a new opportunity to understand how infrastructure
accessibility, quality, and reliability affects telecommuting,
particularly as working from home during the pandemic required high
volumes of bandwidth-intensive video conferencing in order to avoid the
face-to-face contact that could increase the spread of infection. By
first answering questions about internet service resilience, we can also
refine our understanding of how this has contributed to or reduced the
new digital divide, where economic resilience has been dependent upon
the capability to work from home.

The multi-layered digital divides intersect with materials divides and
the economic geography of the UK. Following the regional economic
resilience literature, which underlines the differentiated capacity of
cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}, different places have
different industrial and occupational profiles, and these affect the
aggregated potential capacity of places for telecommuting. Such profiles
are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide
\citep{martin_north_south}. Various studies have illustrated severe
inequalities between the north and the south part of the UK regarding,
for example, skills and human capital, unemployment, productivity and
prosperity \citep{lee2014grim, mccann2020perceptions, dorling2018peak}.
Some scholars have even argued that the UK suffers some of the highest
level of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. Not only all three levels of
digital divides are, to a certain extend, associated with or shaped by
the geography of the UK, but the intersection of the digital and
material divides affects the capacity of places to overcome, at least
partially, the economic effects of the Covid-19 pandemic. Importantly,
this is the first time that digital technologies became an essential
tool for economic resilience for such a great part of the population.

\hypertarget{other-subsection}{%
\subsection{other subsection?}\label{other-subsection}}

\textbf{I THINK WE HAVE ENOUGH OF A LIT REV} - circumstances where work
can be carried out more flexibly in space and time - how relate to
digital divides - capability theory of mobility? - broadband tech stuff?
broadband studies / divides / resilience

Some new papers google recommended to me:

\begin{itemize}
\tightlist
\item
  \url{https://urbanstudies.uva.nl/binaries/content/assets/subsites/centre-for-urban-studies/working-paper-series/wps_43.pdf}
\item
  \url{https://link.springer.com/article/10.1007/s11116-020-10136-6}
\item
  \url{https://www.sciencedirect.com/science/article/pii/S0966692319311305}
\item
  check who cites the above and what they cite
\end{itemize}

\textbf{MORE SOURCES:}

\begin{itemize}
\tightlist
\item
  \url{https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas}
\item
  \href{https://journals.sagepub.com/toc/EPB/current}{EPB commentaries}
\item
  \url{https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home}
\item
  \url{https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity}
\item
  \url{https://www.coronavirusandtheeconomy.com/question/how-will-economic-effects-coronavirus-vary-across-areas-uk}
\item
  \url{https://www.coronavirusandtheeconomy.com/which-parts-uk-have-been-hit-hardest-covid-19-crisis}
\item
  \url{https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas}
\end{itemize}

\hypertarget{methods-and-data}{%
\section{Methods and data}\label{methods-and-data}}

The starting point of our methodological framework is cluster analysis,
which can be defined within the modern machine learning framework as an
unsupervised learning task, which involves partitioning unlabelled
observations into homogeneous groups known as clusters
\citep{montero2014tsclust}. The key idea is that observations within
clusters tend to be more similar than observations between clusters.
Clustering is particularly useful for exploratory studies as it
identifies structures within the data \citep{aghabozorgi2015time}.
Therefore, cluster analysis is a widely used family of techniques in
geography \citep{gordon1977classification, everitt1974cluster}. For
instance, clustering methods is the basis of \emph{geodemographics},
which aims to create small area indicators or typologies of
neighbourhoods based on various and some times diverse variables
\citep{SINGLETON2009289, harris2005geodemographics}. Clustering
techniques have also been employed to solve regionalisation problems
\citep{niesterowicz2016}.

Common characteristic of these studies is the cross-sectional nature of
the data they employ. Indeed, most clustering problems in geography deal
with observations that are fixed in time. However, for this paper we are
interested in creating clusters of local authorities in the UK with
similar temporal signatures of experienced internet speeds over time.
Hence, we deviate from the established geographical clustering tools and
employ time-series clustering methods.

Time-series clustering methods have been developed in order to deal with
clustering problems linked to, for instance, stock or other financial
data, economic, governmental or medical data as well as machine
monitoring
\citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
The main challenge -- and also the difference with cross-sectional
clustering problems -- is data dimensionality given the multiplicity of
data points for every individual object -- local authority in our case
-- included in the data set. Time-series are dynamic data as the value
of the observations change as a function of time
\citep{aghabozorgi2015time}. This high dimensionality leads to (i)
computational and algorithmic challenges regarding handling these data
and building algorithms to perform clustering over long time-series, and
(ii) open questions regarding the choice of similarity measures in order
to cluster similar times series objects together considering the whole
length of the time-series and overcoming issues around noise, outliers
and shifts \citep{lin2004iterative, aghabozorgi2015time}.

Time-series clustering methods utilising the whole length of time-series
can be grouped in three categories. The first -- model-based approaches
-- is based on recovering the underlying model for each time-series and
then applying clustering algorithms on the model parameters of each
time-series \citep{aghabozorgi2015time}. The main criticism is the
cluster accuracy for nearby clusters \citep{mitsa2010temporal}. The
second approach is a based the formation of vectors of features based on
the original time-series. These new data of reduced dimensionality is
then clustered using conventional clustering algorithms.

For this paper we utilise the third category of time-series clustering
methods known as shape-based approaches. These methods match two sperate
time-series objects based on the similarity of their shapes through the
calculation of distances among these two shapes and are better equipped
to capture similarities between short length time-series
\citep{aghabozorgi2015time}. This approach serves best this paper
because (i) we aim to identify clusters of UK local authorities with
similar temporal signatures -- i.e.~shapes -- of experienced internet
speeds and (ii) the length of our time-series is short (see the data
discussion in this section).

Another import element of time-series clustering is the actual
clustering algorithm. Similar to clustering of cross-sectional data, we
can employ partitioning algorithms, which lead to non-overlapping
clusters, hierarchical clustering, which result to a hierarchy of
clusters and fuzzy algorithms, which create overlapping clusters
\citep{sardatime}. Because of the simplicity of the implementation and
the interpretability of the results, we utilise here partitioning
clustering based on the widely used \emph{k-Means} algorithm. This is an
iterative algorithm, which begins with defining the desired number of
clusters \emph{k}. Then, each observation is randomly assigned to a
cluster from the \([1,k]\) space. This is the initial cluster
assignment, which is followed by iterations in order to minimise the
distance between the centroids of the clusters and the observations
assigned to these clusters \citep{james2013introduction}.

There are a number of differences between the above described
application of \emph{k-Means} for cross-sectional data and its
application for times series data. Instead of creating clusters around
centroids, a common approach is to create clusters around
\emph{medoids}, which are representative time-series objects with a
minimal distance to all other cluster objects \citep{sardatime}. Also,
instead of calculating the Euclidean distance between centroids and data
points, more complex distance measures need to be employed in order to
capture the similarity between a time-series object and a medoid. A
common distance measure for shape-based time-series clustering in the
Dynamic Time Warping (DTW). Using its underpinning dynamic programming
algorithm, DTW compares two time-series objects to find the optimum
warping path between them. DTW is widely used in order to overcome
limitations linked to the use of Euclidean distance
\citep{sardatime, berndt1994using, ratanamahatana2004everything}. The
\texttt{R} package \texttt{dtwclust} has been used for the time-series
clustering \citep{dtwclust}.

To assess the quality and reliability of internet across local
authorities in the UK during the time when the population were told to
work from home if at all possible we utilise unique data regarding
individual internet speed tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
This a private company that allows internet users to check their own
broadband upload and download speeds, and stores every speed-check with
timestamp and geolocation information. These data have been used before
to assess digital divides \citep{riddlesden2014broadband} and the impact
of local loop unbundling regulatory processes
\citep{nardotto2015unbundling}. These volunteered geographic data enable
to assess the \emph{experienced} internet speed by users, which may
differ from the \emph{advertised} maximum speeds provided by Internet
Service Providers. We are particularly interested in upload speeds and
the frequency of speed tests. While the former is less associated with
internet-based leisure activities such as video streaming, which involve
downloading large amounts of data, the latter can be linked more with
work-related activities such as uploading documents or two-way
communication and interaction systems. \textbf{ADD MORE. HOW ABOUT SPEED
TESTS?}

The first step in the workflow after dropping some outliers following
\citet{riddlesden2014broadband} is to transform and the individual,
geolocated and time-stamped tests to more meaningful aggregates both in
terms of space and time. Regarding the temporal dimension, we aggregate
all the speed-checks during the \(13\) weeks of March to May inclusive
for weekdays in \(2020\) by each hour of the day and day of the week. As
our research aims to identify the geography of internet service
resilience for work purposes, bank holidays and the hours between
midnight and 6am were also excluded. The composite week time-series thus
comprise 18 hours multiplied by 5 weekdays or 90 time points per series.
We aggregate these data because, we could not follow individuals or
households and connect data points. Also, although this is a large data
set (\(241,088\) individual tests during the working hours of study
period), there are not enough observations for each local and for each
working hour (\(631\) speed test per LAD in average). These time-series
were calculated for each of the 382 Local Authority Districts (LADs) in
the UK, standardised, and then a \emph{k}-means partitioning around
medoids clustering algorithm was applied using DTW. We initially run the
algorithm for \(k = 5, 10, 15 and 20\) and used cluster validity indices
(CVIs) to pick the optimal solution of \(k = 10\). Following
\citet{sardatime} the following CVIs were employed: Silhouette (max),
Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min),
Modified Davies-Bouldin (DB*, min), Dunn (max), COP (min).

We then match the cluster membership to the LAD-level data to identify
the characteristics of each cluster, including number of LADs, the
descriptive statistics of upload speeds in that cluster, and the
temporal profile by hour of the day and day of the composite week. Thus
we aim to answer the first part of our research question: How resilient
are broadband speeds as experienced in different parts of the UK during
a time of extreme demand?

Since the quality and reliability of internet services vary in time and
space due to both supply and demand-side influences, we use a number of
different measures of experienced upload speeds. These include: a) mean,
experienced connection speed, b) standard deviation or the amount of
fluctuation from the mean, and c) the variation in speeds at particular
times of day when working from home is more likely to take place. We
take account of all three measurements in order to describe upload
speeds as fully as possible. The following \textbf{graphs / tables} show
these statistics for the sample as a whole, as well as similar
statistics related to frequency of testing.

\textbf{Data details and some figures, descriptive stats - include whole
sample time profile for 2019 and 2020 frequency of tests run as part of
why we chose to create the time profiles by hour of the day and day of
the week rather than daily over the whole period.}

\textbf{Just looking back, it was partly the frequency of tests, but
that was also by geography, so do we need maps? And I feel there was
something else I'm forgetting\ldots{}}

@@

\begin{align}
Pr(Y_{i}=j) = \frac{exp^{X_{i}\beta_{j}}}{\sum_{i=1}^j exp^{X_{i}\beta_{j}}}
\begin{cases}
    i = 1, 2, ... , N \\t
    j = 1, 2, ... , J
\end{cases}\label{eq1}
\end{align}

The identification strategy is as follows. Based on the outcomes of the
time-series clustering, we identify \(J\) distinct and crisp clusters.
We then regress this cluster membership against a vector \(X_{i}\) of
socio-economic and geographic variables, which we are discussed in
details in \textbf{section XX}. This analysis enables us to provide a
more nuanced understanding of how telecommuting and technology intersect
at a time of extreme demand, and what lessons this time has for a future
where telecommuting is likely to remain a common means of accessing work
and broadband services, as well as infrastructure, must be fit for
purpose.

\hypertarget{results}{%
\section{Results}\label{results}}

>>>>

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{dcolumn}
\def\tightlist{}

@@

\hypertarget{sec:1}{%
\section{Introduction}\label{sec:1}}

During the pandemic, working from home using digital technologies,
whether partially or exclusively, was transformed from a niche means of
accessing work, albeit one that had been on a slow, upward trend, to a
widespread way of life in many countries. The ability to work from home
or telecommute meant millions retained their jobs and, to a varying
extent, maintained productivity during periods of strict lockdowns
around the world. However, this ability has not been evenly distributed
socially or spatially, creating a new type of digital divide. On one
side are those who can work from home, supported by digital
technologies, and have thus been able to enjoy both economic resilience
and greater personal safety. On the other side, previously employed
individuals have been forced to accept furlough or redundancy packages
unless they are part of the cadre of essential workers, who are
potentially at high risk of infection. Whilst the basis for this new
digital divide has been viewed as mainly occupational, here we consider
whether the divide is also technological.

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds, may reinforce or redress the spatial
and social dimensions of the digital division exposed by the pandemic.
To do so, we employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. Our analysis enables
us to better understand how the spatial and social distribution of both
occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand. We will also
consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era, and broadband services and
infrastructure must be fit for purpose. \textbf{LET'S LEAVE IT FOR NOW,
BUT I THINK WE CAN CRYSTALISE MORE THE RQ}

@@

The structure of this paper is as follows. First we review the
literature on telecommuting and digital divides to better understand the
origins of the new digital divide revealed by the pandemic and its
impact on the economic resilience of different places. We then describe
our data and methodology. Our results section first offers a
classification of how internet services vary across the UK local
authorities and then assesses whether these clusters replicate or
repudiate other socio-economic and geographic patterns of economic
resilience.

\hypertarget{sec:2}{%
\section{Literature review}\label{sec:2}}

\hypertarget{sec:2.1}{%
\subsection{From telecommuting to \#WFH}\label{sec:2.1}}

In this analysis, the terms `telecommuting' and `working from home' are
used interchangeably, as most remote labour during the Covid-19 crisis
was carried out in the homes of individual employees rather than any
other location \citep{eurofound2020}. However, it should be noted that
previous research has explored how telecommuting can occur in other
places, including satellite offices or on public transport
\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
also used a variety of definitions to measure the level of telecommuting
within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or
part-time, and those who use digital technologies to work remotely
full-days or part-days
\citep{allen2015effective, bailey2002review, haddad2009examination}. No
matter the definition, the option and capability to telecommute or work
from home has never been equally distributed spatially or
socio-economically any more than different industries and employment
opportunities have. For example, studies from the United States, the
Netherlands, and the UK found that telecommuters are most likely to hold
professional, managerial, and technical occupations where the workforce
is better educated and wealthier, and that there is suppressed demand
among women and part-time workers
\citep{headicar2016move, peters2004employees, singh2013modeling}.
Opportunities for working from home during the current pandemic have
likewise not been equally spread across the workforce.
\citet{NBERw26948} indicated that in the US, managers, educators, as
well as those working in computer-related occupations, finance, and law
can easily work from home, and that occupations with opportunities to
telecommute are associated with higher earnings. This is not the case
for the workforce occupied in more spatially fixed occupations, from
farming, construction and manufacturing to hospitality and care
services. In the US, these occupations tend to be lower-income,
non-white, without a university degree, live in rental accommodation and
lack health insurance \citep{NBERw27085}. Similar trends can be observed
for other countries. For example, \(75\)\% of workers with tertiary
education worked from home in Europe during spring \(2020\), whilst only
\(34\)\% of workers with secondary education and \(14\)\% of those
primary education did so \citep{eurofound2020}.

\hypertarget{sec:2.2}{%
\subsection{Digital divides and economic resilience}\label{sec:2.2}}

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
These returns have been even greater during the Covid-19 crisis, when
those with the ability to telecommute also have the ability to maintain
their employment whilst protecting their health. However, the success of
these arrangements has been dependent upon the first level digital
divide, which is associated with access and quality of internet
connectivity at a time of extremely high demand. \citet{SALEMINK2017360}
provides a systematic review of the pre-pandemic, first level digital
divide in infrastructure quality between urban and rural areas in
various advanced economies. Yet whether this variation in infrastructure
quality affects the spatial footprint of telecommuting has not
previously been measured. There are indications that those who purchase
high speed connections consume more data of all sorts and use their
connections for a variety of purposes \citep{hauge2011consumer}, and
that there is a correlation between access to internet services and a
reduction in household transport spend \citep{bris2017ict}. Whether the
implication is more internet use and less travel because of increased
telecommuting, these studies suggest that better internet services
enable households to make savings and efficiencies, an example of the
first level digital divide reinforcing the third level. Yet did the
purchase of high speed connections and increased internet access also
prepare households for long-term home-working, enforced by government
restrictions? The extreme demand during the pandemic provides a new
opportunity to understand how infrastructure accessibility, quality, and
reliability affects telecommuting, particularly in light of the high
volumes of bandwidth-intensive video conferencing required in order to
avoid the face-to-face contact that could increase the spread of
infection. We seek to answer how internet service resilience contributes
to or reduces economic resilience when the latter is dependent upon the
capability to work from home, and thus improve our understanding of this
new digital divide. \textbf{RELATED TO YOUR POINT ABOVE, WE MAY CONCLUDE
THIS IS NOT A NEW DIGITAL DIVIDE, BUT THAT IS STILL ANSWERING A USEFUL
QUESTION - MIGHT END UP REWORDING AS INTERACTION BETWEEN DIGITAL
DIVIDES? ET: OR REINFORCING THE 3RD LEVEL DIGITAL DIVIDE}

The multi-layered digital divides intersect with material divides and
the economic geography of the UK. Following the regional economic
resilience literature, which underlines the differentiated capacity of
cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}, different places have
different industrial and occupational profiles, and these affect the
aggregated potential capacity of places for telecommuting. Such profiles
are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide
\citep{martin_north_south}. Various studies have illustrated severe
inequalities between the north and the south part of the UK regarding,
for example, skills and human capital, unemployment, productivity and
prosperity \citep{lee2014grim, mccann2020perceptions, dorling2018peak}.
Some scholars have even argued that the UK suffers some of the highest
level of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
digital divides associated to a certain extent with or shaped by the
geography of the UK, but the intersection of the digital and material
divides affects the capacity of places to overcome some of the economic
effects of the Covid-19 pandemic. Importantly, this is the first time
that digital technologies became an essential tool for economic
resilience for such a great part of the population.

\hypertarget{sec:3}{%
\section{Methods and data}\label{sec:3}}

The starting point of our methodological framework is cluster analysis,
which can be defined within the modern machine learning framework as an
unsupervised learning task, partitioning unlabelled observations into
homogeneous groups known as clusters \citep{montero2014tsclust}. The key
idea is that observations within clusters tend to be more similar than
observations between clusters. Clustering is particularly useful for
exploratory studies as it identifies structures within the data
\citep{aghabozorgi2015time}. Therefore, cluster analysis is a widely
used family of techniques in geography
\citep{gordon1977classification, everitt1974cluster}. For instance,
clustering methods are the basis of \emph{geodemographics}, a research
domain which aims to create small area indicators or typologies of
neighbourhoods based on various and sometimes diverse variables
\citep{SINGLETON2009289, harris2005geodemographics}. Clustering
techniques have also been employed to solve \emph{regionalisation}
problems \citep{niesterowicz2016}.

Common characteristics of these studies are the cross-sectional nature
of the data they employ. Indeed, most clustering problems in geography
deal with observations that are fixed in time. However, for this paper
we are interested in internet speeds, which vary over time and,
therefore, create clusters of local authorities in the UK with similar
temporal signatures of experienced internet speeds. Hence, we deviate
from the established geographical clustering tools and employ
time-series clustering methods.

Time-series clustering methods have been developed in order to deal with
clustering problems linked to, for instance, stock or other financial
data, economic, governmental or medical data as well as machine
monitoring
\citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
The main challenge -- and also the difference with cross-sectional
clustering problems -- is data dimensionality given the multiplicity of
data points for every individual object -- local authorities in our case
-- included in the data set. Time-series are dynamic data as the value
of the observations change as a function of time
\citep{aghabozorgi2015time}. This high dimensionality leads to (i)
computational and algorithmic challenges regarding handling these data
and building algorithms to perform clustering over long time-series, and
(ii) open questions regarding the choice of similarity measures in order
to cluster similar times series objects together considering the whole
length of the time-series and overcoming issues around noise, outliers
and shifts \citep{lin2004iterative, aghabozorgi2015time}.

Time-series clustering methods utilising the whole length of time-series
can be grouped in three categories. The first -- model-based approaches
-- is based on recovering the underlying model for each time-series and
then applying clustering algorithms on the model parameters of each
time-series \citep{aghabozorgi2015time}. The main criticism is the
cluster accuracy for nearby clusters \citep{mitsa2010temporal}. The
second approach is based the formation of vectors of features derived
from the original time-series. These new data of reduced dimensionality
are then clustered using conventional clustering algorithms.

For this paper we utilise the third category of time-series clustering
methods known as shape-based approaches. These methods match two sperate
time-series objects based on the similarity of their shapes through the
calculation of distances between the shapes, and are thus better
equipped to capture similarities between short length time-series
\citep{aghabozorgi2015time}. This approach serves best this paper
because (i) we identify clusters of UK local authorities with similar
temporal signatures -- i.e.~shapes -- of experienced internet speeds and
(ii) the length of our time-series is short (see the data discussion in
this section).

Another import element of time-series clustering is the actual
clustering algorithm. Similar to the clustering of cross-sectional data,
we can employ partitioning algorithms, which lead to non-overlapping
clusters, hierarchical clustering, which classifies clusters at
different levels, and fuzzy algorithms, which create overlapping
clusters \citep{sardatime}. Because of the simplicity of the
implementation and the interpretability of the results, we utilise here
partitioning clustering based on the widely used \emph{k-Means}
algorithm. This is an iterative algorithm, which begins with defining
the desired number of clusters \emph{k}. Then each observation is
randomly assigned to a cluster from the \([1,k]\) space. This initial
cluster assignment is followed by iterations in order to minimise the
distance between the centroids of the clusters and the observations
assigned to these clusters \citep{james2013introduction}.

There are a number of differences between the above described
application of \emph{k-Means} for cross-sectional data and its
application for times series data. Instead of creating clusters around
centroids, a common approach is to create clusters around
\emph{medoids}, which are representative time-series objects with a
minimal distance to all other cluster objects \citep{sardatime}. Also,
instead of calculating the Euclidean distance between centroids and data
points, more complex distance measures need to be employed in order to
capture the similarity between a time-series object and a medoid. A
common distance measure for shape-based time-series clustering is
Dynamic Time Warping (DTW). Using its underpinning dynamic programming
algorithm, DTW compares two time-series objects to find the optimum
warping path between them. DTW is widely used in order to overcome
limitations linked to the use of Euclidean distance
\citep{sardatime, berndt1994using, ratanamahatana2004everything}. The
\texttt{R} package \texttt{dtwclust} has been used for the time-series
clustering \citep{dtwclust}.

To assess the quality and reliability of internet across local
authorities in the UK during the time when the population were told to
work from home if at all possible we utilise unique data comprising
individual internet speed tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
This a private company that allows internet users to check their own
broadband upload and download speeds, and stores every speed-check with
timestamp and geolocation information. These data have been used before
to assess digital divides \citep{riddlesden2014broadband} and the impact
of local loop unbundling regulatory processes
\citep{nardotto2015unbundling}. By using volunteered geographic data, we
are able to assess the \emph{experienced} internet speed by users, which
may differ from the \emph{advertised} maximum speeds of Internet Service
Providers (ISPs). We are particularly interested in upload speeds and
the frequency of speed tests over the period from March to May 2020, as
government statements indicate this is when UK workers were told to work
from home if at all possible \textbf{{[}need ref?{]}}. Average upload
speeds are slower than average download speeds, at \(9.3\)Mb/s mean
upload speed for the whole sample during the period of interest,
compared to \(29.6\)Mb/s for download speeds, but they are also less
associated with internet-based, high-demand, leisure activities such as
video streaming. Therefore, upload speeds are more relevant to
work-related activities such as uploading documents or two-way audio,
video, and text-based communication systems. Meanwhile, the frequency of
speed tests was important in identifying the temporal profile which woud
give us most insight into experienced internet service and resilience,
and provide an indication of the volume of experience over particular
units of time.

The first step in the workflow after dropping some outliers following
\citet{riddlesden2014broadband} was to transform the individual,
geolocated and time-stamped tests to more meaningful aggregates both in
terms of space and time. The frequency of testing indicates that whilst
there is an overall trend of increased testing \textbf{or increased
experience of internet service NOT SURE WHAT THIS MEAND} from March to
April and then a slight reduction from April to May, this trend masks
substantial variation by not only the day of the week, but also time of
day, as can be seen in Figure \ref{test2020}. Thus, a daily aggregation
of upload speeds would mask the variation in experience over the course
of each weekday. Furthermore, the importance of this variation is
highlighted by a comparison with the same period in \(2019\), as in
Figure \ref{test2019}, when the volume of testing and thus of experience
of internet service quality was much more tied to the evening leisure
peak and presumably to download speeds. Since the increase in testing
during the working day in \(2020\) is an indication that users have a
greater perception of the variable quality of internet service,
particularly during a new morning peak of testing for service
reliability, we decided to include a measure of hourly variation in our
temporal profiles. However, although this is a large data set
(\(241,088\) individual tests during the working hours of study period),
there are not enough observations for each Local Authority District
(LAD) and for each working hour of each working day (\(631\) speed tests
per LAD on average) to profile speeds at that level of detail.
Therefore, we aggregate all the speed-checks during the \(13\) weeks of
March to May inclusive for weekdays in \(2020\) by each hour of the day
and day of the week. As our research aims to identify the geography of
internet service resilience for work purposes, bank holidays and the
hours between midnight and 6:00 were excluded, as well as weekend days.
The composite week time-series thus comprise \(18\) hours multiplied by
\(5\) weekdays or \(90\) time points per series. We also aggregate these
data spatially because we could not follow individuals or households and
connect data points. The time-series were calculated for each of the
\(382\) LADs in the UK, standardised, and then a \emph{k}-means
partitioning around medoids clustering algorithm was applied using DTW.
We initially run the algorithm for \(k \in \mathbb{N} \bigcap [5,15]\)
and used cluster validity indices (CVIs) to pick the optimal solution of
\(k = 13\). Following \citet{sardatime} the majority vote for the
following CVIs was used: Silhouette (max), Score function (max),
Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin
(DB*, min), Dunn (max), COP (min).

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth,height=0.4\textheight]{figures/time.var.plot2020.png}
\caption{Speed tests over time, 2020 \label{test2020}}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth,height=0.4\textheight]{figures/time.var.plot2019.png}
\caption{Speed tests over time, 2019 \label{test2019}}
\end{figure}

We then review the characteristics of each cluster, including number of
LADs, and the temporal profile of upload speed by hour of the day and
day of the composite week. Since the quality and reliability of internet
services vary in time and space due to both supply and demand-side
influences, we use a number of different measures to then describe
experienced upload speeds per cluster. These include: a) mean,
experienced connection speed, b) standard deviation or the amount of
fluctuation from the mean, and c) the variation in speeds at particular
times of day when working from home is more likely to take place. We
take account of all three measurements in our descriptives statistics of
upload speeds in order to determine how resilient broadband speeds are
as experienced in different parts of the UK during a time of extreme
demand.

\textbf{More data details and some figures, descriptive stats}

@@

\begin{align}
Pr(Y_{i}=j) = \frac{exp^{X_{i}\beta_{j}}}{\sum_{i=1}^j exp^{X_{i}\beta_{j}}}
\begin{cases}
    i = 1, 2, ... , N \\  
    j = 1, 2, ... , J
\end{cases}\label{eq1}
\end{align}

The identification strategy is as follows. Based on the outcomes of the
time-series clustering, we identify \(J\) distinct and crisp clusters.
We then regress this cluster membership against a vector \(X_{i}\) of
socio-economic and geographic variables, which are discussed in detail
in Section \ref{#sec:4.2}. This analysis enables us to provide a more
nuanced understanding of how telecommuting and technology intersect at a
time of extreme demand, and what lessons this time has for a future
where telecommuting is likely to remain a common means of accessing work
and broadband services, as well as infrastructure, must be fit for
purpose.

\hypertarget{sec:4}{%
\section{Results}\label{sec:4}}

\hypertarget{sec:4.1}{%
\subsection{Upload Clusters / cluster description}\label{sec:4.1}}
